Implementation of a lattice Boltzmann method for immiscible multiphase flow simulations using the level set method

### 2015-11-11 Paper/paper.tex
Results angefangen. Parameterangaben und Plots fehlen noch.

\begin{figure}[h!]
	\flushright
	\includegraphics[width=12cm,natwidth=1690,natheight=1174]{skizze.png}
	\caption{Two fluid domains $\Omega_i$ and the interface $\Gamma$ in between}
	\label{intro_phases}
\end{figure}

@@

\subsection{The Lattice Boltzmann Method}
To solve the incompressible Navier-Stokes equations, we use the lattice Boltzmann method. This method describes the evolution of the particle density $f(\vec{x},\vec{v},t)$ in phase space, with a $(\vec{x},\vec{v})$ as the phase space variables and $t$ as time. The LBM algorithm discretizes with a regular grid in space and with restricted number of velocities adapted to this grid \cite{LBM1,LBM2,LBM3,LBM4}. We use the D2Q9 model in 2D, which has 9 velocity vectors on a plane grid with unit spacing including one zero velocity (Fig. \ref{lbmc_i}). 
\begin{figure}
	\hfill\includegraphics[width=6cm,natwidth=403,natheight=409]{LBMc_i.png}\hspace*{\fill}
	\caption{The vectors of one cell in LBM}
	\label{lbmc_i}
\end{figure}
The vectors are given by the columns of the matrix
\begin{equation}
	c = 
	\begin{bmatrix}
	0 & 0 & 1 & 1 &  1 &  0 & -1 & -1 & -1 \\
	0 & 1 & 1 & 0 & -1 & -1 & -1 &  0 &  1
	\end{bmatrix}
	\text{.}
\end{equation}
The corresponding particle distributions are denoted by $f_i(\vec{x},t) = f(\vec{x},\vec{c_i},t)$. We get the density and velocity from the particle distribution functions with:
\begin{equation}
	\rho(\vec{x},t) = \sum_{i=0}^{8} f_i(\vec{x},t) \text{, } 
	\vec{u} = \sum_{i=0}^{8} f_i(\vec{x},t)\vec{c_i} \text{.}
\end{equation}
For the collision step we use the Bhatnagar-Gross-Krook (BGK) approximation \cite{BGK}
\begin{equation}
	f_i(\vec{x}+\vec{c_i},t+1) = f_i(\vec{x},t) - \frac1\tau(f_i-f_i^{eq}) + G_i \text{.}
\end{equation}
The parameter $\tau$ is the relaxation parameter for the BGK collision operator and controls the kinematic viscosity $\nu = \frac16(2\tau - 1)$. $G_i$ is an additional force like gravity or a magnetic field. In our case we set $G_i = 0$. Furthermore, we use the equilibrium distribution
\begin{equation}
	f_i^{eq}(\rho,\vec{u}) = f_i^*(\rho + 3\vec{c_i} \cdot \vec{u} + \frac92(\vec{c_i} \cdot \vec{u})^2 - \frac32{|\vec{u}|}^2)
\end{equation}
with the corresponding D2Q9 weight factors
$$
f_i^* = \begin{cases} \frac49 \text{,} & i = 0 \\ \frac19 \text{,} & i = 1,3,5,7 \\ \frac{1}{36} \text{,} &  i = 2,4,6,8 \end{cases} \text{.}
$$

The LBM algorithm consists out of the following steps:
\begin{itemize}
	\item[1. ] Collision step: $f_i^+ = f_i - \frac1\tau(f_i - f_i ^{eq}) + G_i$
	\item[2. ] Propagation step: $f_i(\vec{x}+\vec{c_i},t+1) = f_i^+(x,t)$, for interior nodes
	\item[3. ] Boundary conditions: $f_i(\vec{x}+\vec{c_i},t+1) = f_i(\vec{x}+\vec{c_i},t)$, if $\vec{x} \notin \Omega$
\end{itemize}

@@

We can rewrite the formula in the form

@@

Figure \ref{refill} shows the cells we use for the interpolation. In \cite{Thoemmes} the velocity is interpolated by using the velocity of the interface. The Ian-Mitchell toolbox does not provide a function to access this value. Therefore, we interpolate the velocity in the same way as we did for the density:

@@

\section{something new what we did / Results}
Present our test case (I think one would be enough) and describe the parameters we have studied. Give graphs that show their impact on accuracy.

@@

>>>>

\begin{figure}[h!]
	\flushright
	\includegraphics[width=11cm,natwidth=1690,natheight=1174]{skizze.png}
	\caption{Two fluid domains $\Omega_i$ and the interface $\Gamma$ in between}
	\label{intro_phases}
\end{figure}

@@

\subsection{The Lattice Boltzmann Method}
To solve the incompressible Navier-Stokes equations, we use the lattice Boltzmann method. This method describes the evolution of the particle density $f(\vec{x},\vec{v},t)$ in phase space, with $(\vec{x},\vec{v})$ as the phase space variables and $t$ as time. The LBM algorithm discretizes with a regular grid in space and with restricted number of velocities adapted to this grid \cite{LBM1,LBM2,LBM3,LBM4}. We use the D2Q9 model in 2D, which has 9 velocity vectors on a plane grid with unit spacing including one zero velocity (Fig. \ref{lbmc_i}). 
\begin{figure}
	\hfill\includegraphics[width=6cm,natwidth=403,natheight=409]{LBMc_i.png}\hspace*{\fill}
	\caption{The vectors of one cell in LBM}
	\label{lbmc_i}
\end{figure}
The vectors are given by the columns of the matrix
\begin{equation}
	c = 
	\begin{bmatrix}
	0 & 0 & 1 & 1 &  1 &  0 & -1 & -1 & -1 \\
	0 & 1 & 1 & 0 & -1 & -1 & -1 &  0 &  1
	\end{bmatrix}
	\text{.}
\end{equation}
The corresponding particle distributions are denoted by $f_i(\vec{x},t) = f(\vec{x},\vec{c_i},t)$. We get the density and velocity from the particle distribution functions with:
\begin{equation}
	\rho(\vec{x},t) = \sum_{i=0}^{8} f_i(\vec{x},t) \text{, } 
	\vec{u} = \sum_{i=0}^{8} f_i(\vec{x},t)\vec{c_i} \text{.}
\end{equation}
For the collision step we use the Bhatnagar-Gross-Krook (BGK) approximation \cite{BGK}
\begin{equation}
	f_i(\vec{x}+\vec{c_i},t+1) = f_i(\vec{x},t) - \frac1\tau(f_i-f_i^{eq}) + G_i \text{.}
\end{equation}
The parameter $\tau$ is the relaxation parameter for the BGK collision operator and controls the kinematic viscosity $\nu = \frac16(2\tau - 1)$. $G_i$ is an additional force like gravity or a magnetic field. In our case we set $G_i = 0$. Furthermore, we use the equilibrium distribution
\begin{equation}
	f_i^{eq}(\rho,\vec{u}) = f_i^*(\rho + 3\vec{c_i} \cdot \vec{u} + \frac92(\vec{c_i} \cdot \vec{u})^2 - \frac32{|\vec{u}|}^2)
\end{equation}
with the corresponding D2Q9 weight factors
$$
f_i^* = \begin{cases} \frac49 \text{,} & i = 0 \\ \frac19 \text{,} & i = 1,3,5,7 \\ \frac{1}{36} \text{,} &  i = 2,4,6,8 \end{cases} \text{.}
$$

One time step in the LBM algorithm is achieved by the following sub-steps:
\begin{itemize}
	\item[1. ] Collision step: $f_i^+ = f_i - \frac1\tau(f_i - f_i ^{eq}) + G_i$
	\item[2. ] Propagation step: $f_i(\vec{x}+\vec{c_i},t+1) = f_i^+(x,t)$, for interior nodes
	\item[3. ] Boundary conditions: $f_i(\vec{x}+\vec{c_i},t+1) = f_i(\vec{x}+\vec{c_i},t)$, if $\vec{x} \notin \Omega$
\end{itemize}

@@

We can rewrite $R_i$ in the form

@@

Figure \ref{refill} shows the cells we use for the interpolation. In \cite{Thoemmes} the velocity is interpolated by using the velocity of the interface. The Ian-Mitchell toolbox does not provide a function to access this value. Therefore, we follow the formula given in \cite{Lallemand} that is basically the same we use for density:

@@

\section{Results and Examples}
To verify the correctness of our implementation we firstly created a Couette flow and compared our results with the one we get from the analytic solution. Secondly, we used the same LBM setup but with a different fluid distribution to check the behavior of our implementation for more complex problems.

\subsection{Couette channel flow}
The Couette channel flow consists of a fixed lower wall at $y=0$ and an upper wall at $y=1$ that is moving with a constant velocity. At these walls we use no-slip boundary conditions. To create a channel flow, we use periodic boundary conditions for the boundaries at the inflow $x=0$ and the outflow $x=1$. The two fluids are distributed in a way that the interface is parallel to our walls.
%TODO: give information about the parameter
%TODO: show graph with results

For the Couette channel flow exists an analytic solution that can be derived from the Navier-Stokes equations
$$
v(x) = \begin{cases} v_0 (\frac{\mu_1}{\mu_1+\mu_2}x + \frac{\mu_2}{\mu_1+\mu_2}) \text{,} & x \ge 0 \\ v_0 (\frac{\mu_2}{\mu_1+\mu_2}x + \frac{\mu_2}{\mu_1+\mu_2}) \text{,} & x < 0 \text{,}\end{cases}
$$
where $\mu_1$ denotes the dynamic viscosity in the lower fluid and $\mu_2$ the viscosity in upper fluid.

\subsection{Bubble tracing}
To verify the correctness of our implementation also for more complex problems, we changed the fluid distribution in our Couette channel flow such that one fluid is a bubble within the other fluid. We set all material parameters of the fluids to the same values. Thus one would expect that we get the same flow field as for only a single fluid and the interface merely serves as a tracer. FIGURE ?? shows how the bubble that is initialized as circle changes to an elliptic shape caused by the flow field within the channel.

%TODO: name the settings we used
%TODO: add some nice pictures

@@

\bibitem{Lallemand} {\sc P. Lallemand, L.-S. Luo}, {\em Lattice Boltzmann method for moving boundaries}, J. Comput. Phys. 184 (2003) 406-421
